HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-06-29 scnn.tex
figref, tabref, eqnref, secref

\newcommand{\nati}[1]{{\color[rgb]{.1,.6,.1}{#1}}}

\newcommand{\todo}[1]{{\color[rgb]{.6,.1,.6}{#1}}}

\newcommand{\assign}[1]{{\color[rgb]{.8,.5,.8}{Assigned: #1 }}}

@@

The paper is organised as follows.
In Section~\ref{sec:related}, we describe existing approaches to convolutions on the sphere, graphs, and manifolds.
In Section~\ref{sec:method}, we show the way to construct a graph using Healpix sampling and define the convolution operation.
The comparison between the spherical CNN and the benchmark method using the weak lensing mass map classification is presented in Section~\ref{sec:experiments}.
We conclude in Section~\ref{sec:conclusion}.

@@

\subsection{Graph creation}
In the classical 2-dimenstional convolution, each pixel is connected with the
same weight to its 4 closest neighbors. We use a weighted undirected graph to
generalize this idea to the ubiquitous HealPix
sampling~\citep{gorski2005healpix}.  Each pixel is represented by a node
(vertex) connected to his $8$ or $7$ closest neighbors.\footnote{For some
pixels, the $8^{th}$ nearest neighbor is not well defined.} Given the set of
nearest neighbors, we define the weight matrix $W$ using the following scheme
\begin{equation}
W[i,j]=\begin{cases}
e^{-\frac{\|x_i-x_j\|_2^2}{\sigma^2}} & \text{if pixels $i$ and $j$ are connected, and}\\
0 & \text{otherwise.}\\
\end{cases}
\end{equation}
Here $x_i$ is a 3-dimensional vector encoding the coordinate of the pixels $i$
on the sphere and $\sigma$ is the mean of $\|x_i-x_j\|_2$ over all connected
pixels $i$ and $j$. This weighting scheme is important as the distances between
pixels varies due to the small sampling irregularities. Some of the
irregularities have been highlighted in the graph displayed in
Figure~\ref{fig:healpix_graph_4}.

@@

As shown in Figure~\ref{fig:graph_harmonics}, it turns out that using the graph
construction described above, the graph Fourier harmonics resemble the
classical spherical harmonics giving us an indication that the graph is able to
capture the spherical properties of the HealPix sampling. This topic is further discusses in Appendix~\ref{sec:spherical_harmonics}.

@@

Graph convolution is defined by generalizing the concept that convolution is a
multiplication in the spectral domain. Given a convolution kernel
$h:\mathbb{R}_+\rightarrow\mathbb{R}$, the convolution of a signal $\b{f}$
is defined as
\begin{equation} \label{def:graph_convolution}
h(L)\b{f} = \U h(\bLambda) \U^* \b{f},
\end{equation}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii}=h(\bLambda_{ii})$.
Equation~\ref{def:graph_convolution} contains three parts: a) $\U^* \b{f}$, the
Fourier transform of $\b{f}$; b) $h(\bLambda)$ the multiplication with the Fourier
kernel $h(\blambda)$; and c) $\U$ the inverse Fourier transform.

@@

\paragraph{Example:}
Let us consider the heat diffusion problem
\begin{equation} \label{eq:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition
$\b{f}(0)$, the solution of~\ref{eq:heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U^* \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation}
which is a convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$ that
are one generalization of the squared frequencies on the graph. Hence the kernel $K_t$ is also a generalization of the
Gaussian kernel on the sphere. In Figure~\ref{fig:gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various
time $t$. In  Figure~\ref{fig:gaussian_filters_visualization} of
Appendix~\ref{app:filter_visualization}, we show different visualizations of the
convolution kernel $K_t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(Figure~\ref{fig:gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parameterization, well approximate it.

@@

\subsection{Coarsening}
The HealPix sampling presents the advantage to have a natural coarsening method
based on its construction. Each Nside level divides each pixel into 4 subpixels.
We define the pooling process as the inverse operation grouping the 4 subpixels.
Figure~\ref{fig:pooling} illustrates the process.
\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/pooling.pdf}
\caption{Pooling illustration for 2 levels using some part of the mollview projection.
The final patch represents 1/12 of the sphere.}
\label{fig:pooling}
\end{figure}

@@

We make full sky N-body simulations for two parameter set for $\Lambda \rm{CDM} $ cosmological model: model 1 ($\Omega_m=0.31, \sigma_8=0.82$) and model 2 ($\Omega_m=0.26, \sigma_8=0.91$), where $\Omega_m$ is the matter density in the universe and $\sigma_8$ is the normalisation of the matter power spectrum.
Other parameter parameters were set to $h=0.7$ (MORE PARAMS HERE - CHECK WITH RAPHAEL).
These parameters were chosen to have the same spherical harmonic power spectrum for $\ell<1000$.
That means that it is not possible to distinguish between these cosmological models using these maps if the this range of scales is used.
We found that for $\ell>1000$ the differences in power spectrum is $~5\%$.
To remove this information, we additionally smooth the spherical maps with a Gaussian kernel of radius $3$ arcmins. The resulting PSD are displayed in Figure~\ref{fig:psd_sigma3}.

@@

As a pre-processing step, we first remove the mean of each sample and we apply a spherical Gaussian smoothing kernel with a radius of $1$ arcmin.
Out of the 60 simulations, 20 are kept for the test set.
% The training and validation set are created later on.
Figure \ref{fig:map_sample} shows the full sky simulations and a zoom region for model 1 (top) and model 2 (bottom).
Initial conditions for these simulations were the same, so the differences in structures can only be attributed to different cosmological parameters used to evolve the particle distribution.

@@

Leveraging the properties of the HealPix sampling, we can split each maps
into $12*o^2$ ($o=1,2,4,\dots$) samples that span a smaller part of the sphere. 
The area of the pixels for an Nside of $o$ correspond to the different split of the data.
As shown in Figure~\ref{fig:subpart_sphere}, we used $o=1,2,4$ as the resulting
samples are large enough to suffer from the effects of the spherical geometry. We
decided to only report simulation results for this specific setting as spherical
cosmological data usually does not span the full sphere. Nevertheless, our code
includes an example using the full sphere.

@@

We present in Figure~\ref{fig:results} the results for 5 different noise
levels and $3$ different data sizes.  We observe that the PSD features, i.e.
one of the most used technique for model discrimination in cosmology is unable
to perform the desired task. The main reason is that the two different model
have very similar PSDs (see Figure~\ref{fig:psd_sigma3}).   As a result, one
need to use other statistics to solve the problem. In this case, we use
histogram features and observe a significant improvement over the PSD.  The
SCNN present the advantage to learn automatically specific features for the
task and also the usual Deep Learning performance bump.

@@

\section{Spherical harmonics VS graph Fourier eigenvectors}
Figure~\ref{fig:graph_harmonics} displays the first 16 eigenvectors of the
graph Fourier basis used for the convolution. While their is a clear
resemblance with the spherical harmonic, the small irregularities in the
sampling (see~\ref{fig:healpix_graph_4}) have two important effects on the
graph Fourier eigenvectors.  First, some graph eigenvectors will be
localized~\cite{perraudin2018global}, i.e: they will span only a small subpart
of the sphere.  Second, as Nside increase toward infinity and hence the number
of pixel $n\rightarrow \infty$, we are still unsure if the graph eigenvectors
converge toward the spherical harmonics. We believe that it might be the case
for a different graph construction (fully connected) and that a proof could be
built on top of the work of~\cite{belkin2007convergence}. This is obviously
out of the scope of this contribution.

@@

\section{Filter visualization}
\label{app:filter_visualization}

In Figure~\ref{fig:gaussian_filters_visualization}, we present three different
ways to plot the graph convolution kernel $K_t(x)=e^{-\tau t x}$. First, they can
be plotted in the graph spectral domain. In this case, we simply evaluate $K_t$
at the graph eigenvalues $\text{diag}(\bLambda)$. Second, we can convolve a
delta on the sphere and plot a gnomonic view of the results. Third, since we are
working on the sphere, we can only plot the section of the convoluted delta. In
Figure~\ref{fig:index_section}, we display the selected indexes for the section
plotting. Note that, because of the small irregularities in the HealPix
sampling, the second and the third methods are likely to change depending on the
chosen delta.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/index_plotting_order20_nside16.pdf}
\caption{Indexes selected for the section ploting of Figure~\ref{fig:gaussian_filters_visualization} middle. The delta was placed on the yellow node.}
\label{fig:index_section}
\end{figure}

@@

When comparing the network, we want to ensure that the SVM classifier has
access to the same amount of data that the network, i.e, potentially an
infinite number of sample. Hence, we fit different linear SVM classifier using
various amount of training set size until we oberve experimentally that
increasing the amount of data does not improve the classifier. The validation
set remain the same and is used for cross-validation to tune the L2
penalization parameter of SVM. Eventually, we evaluate the final classifier on
the testing set. An example of the evolution of the training and valdication
error is show in Figure~\ref{fig:hist_error_evolution}.

>>>>

\newcommand{\nati}[1]{{\color[rgb]{.1,.6,.1}{#1}}}
\newcommand{\todo}[1]{{\color[rgb]{.6,.1,.6}{#1}}}
\newcommand{\assign}[1]{{\color[rgb]{.8,.5,.8}{Assigned: #1 }}}

\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
\newcommand{\tabref}[1]{Table~\ref{tab:#1}}
\newcommand{\secref}[1]{Section~\ref{sec:#1}}
%\newcommand{\secref}[1]{\S\ref{sec:#1}}
\newcommand{\eqnref}[1]{(\ref{eqn:#1})}

@@

The paper is organised as follows.
In \secref{related}, we describe existing approaches to convolutions on spheres, graphs, and manifolds.
\secref{method} exposes the construction of a graph given the Healpix sampling of a sphere and defines the convolution operation.
Finally, \secref{experiments} presents our experimental results on the weak lensing mass map classification problem. There we compare our method to two standard benchmarks and show its superiority.
%We conclude in \secref{conclusion}.

@@

\subsection{Graph creation}
In the classical 2-dimenstional convolution, each pixel is connected with the
same weight to its 4 closest neighbors. We use a weighted undirected graph to
generalize this idea to the ubiquitous HealPix
sampling~\citep{gorski2005healpix}.  Each pixel is represented by a node
(vertex) connected to his $8$ or $7$ closest neighbors.\footnote{For some
pixels, the $8^{th}$ nearest neighbor is not well defined.} Given the set of
nearest neighbors, we define the weight matrix $W$ using the following scheme
\begin{equation}
W[i,j]=\begin{cases}
e^{-\frac{\|x_i-x_j\|_2^2}{\sigma^2}} & \text{if pixels $i$ and $j$ are connected, and}\\
0 & \text{otherwise.}\\
\end{cases}
\end{equation}
Here $x_i$ is a 3-dimensional vector encoding the coordinate of the pixels $i$
on the sphere and $\sigma$ is the mean of $\|x_i-x_j\|_2$ over all connected
pixels $i$ and $j$. This weighting scheme is important as the distances between
pixels varies due to the small sampling irregularities. Some of the
irregularities have been highlighted in the graph displayed in
\figref{healpix_graph_4}.

@@

As shown in \figref{graph_harmonics}, it turns out that using the graph
construction described above, the graph Fourier harmonics resemble the
classical spherical harmonics giving us an indication that the graph is able to
capture the spherical properties of the HealPix sampling. This topic is further discusses in Appendix~\ref{sec:spherical_harmonics}.

@@

Graph convolution is defined by generalizing the concept that convolution is a
multiplication in the spectral domain. Given a convolution kernel
$h:\mathbb{R}_+\rightarrow\mathbb{R}$, the convolution of a signal $\b{f}$
is defined as
\begin{equation} \label{eqn:graph_convolution}
h(L)\b{f} = \U h(\bLambda) \U^* \b{f},
\end{equation}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii}=h(\bLambda_{ii})$.
Equation~\eqnref{graph_convolution} contains three parts: a) $\U^* \b{f}$, the
Fourier transform of $\b{f}$; b) $h(\bLambda)$ the multiplication with the Fourier
kernel $h(\blambda)$; and c) $\U$ the inverse Fourier transform.

@@

\paragraph{Example:}
Let us consider the heat diffusion problem
\begin{equation} \label{eqn:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U^* \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation}
which is a convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$ that
are one generalization of the squared frequencies on the graph. Hence the kernel $K_t$ is also a generalization of the
Gaussian kernel on the sphere. In \figref{gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various
time $t$. In \figref{gaussian_filters_visualization} of
Appendix~\ref{sec:filter_visualization}, we show different visualizations of the
convolution kernel $K_t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(\figref{gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parameterization, well approximate it.

@@

\subsection{Coarsening}
The HealPix sampling presents the advantage to have a natural coarsening method
based on its construction. Each Nside level divides each pixel into 4 subpixels.
We define the pooling process as the inverse operation grouping the 4 subpixels.
\figref{pooling} illustrates the process.
\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/pooling.pdf}
\caption{Pooling illustration for 2 levels using some part of the mollview projection.
The final patch represents 1/12 of the sphere.}
\label{fig:pooling}
\end{figure}

@@

We make full sky N-body simulations for two parameter set for $\Lambda \rm{CDM} $ cosmological model: model 1 ($\Omega_m=0.31, \sigma_8=0.82$) and model 2 ($\Omega_m=0.26, \sigma_8=0.91$), where $\Omega_m$ is the matter density in the universe and $\sigma_8$ is the normalisation of the matter power spectrum.
Other parameter parameters were set to $h=0.7$ (MORE PARAMS HERE - CHECK WITH RAPHAEL).
These parameters were chosen to have the same spherical harmonic power spectrum for $\ell<1000$.
That means that it is not possible to distinguish between these cosmological models using these maps if the this range of scales is used.
We found that for $\ell>1000$ the differences in power spectrum is $~5\%$.
To remove this information, we additionally smooth the spherical maps with a Gaussian kernel of radius $3$ arcmins. The resulting PSD are displayed in \figref{psd_sigma3}.

@@

As a pre-processing step, we first remove the mean of each sample and we apply a spherical Gaussian smoothing kernel with a radius of $1$ arcmin.
Out of the 60 simulations, 20 are kept for the test set.
% The training and validation set are created later on.
\figref{map_sample} shows the full sky simulations and a zoom region for model 1 (top) and model 2 (bottom).
Initial conditions for these simulations were the same, so the differences in structures can only be attributed to different cosmological parameters used to evolve the particle distribution.

@@

Leveraging the properties of the HealPix sampling, we can split each maps
into $12*o^2$ ($o=1,2,4,\dots$) samples that span a smaller part of the sphere. 
The area of the pixels for an Nside of $o$ correspond to the different split of the data.
As shown in \figref{subpart_sphere}, we used $o=1,2,4$ as the resulting
samples are large enough to suffer from the effects of the spherical geometry. We
decided to only report simulation results for this specific setting as spherical
cosmological data usually does not span the full sphere. Nevertheless, our code
includes an example using the full sphere.

@@

We present in \figref{results} the results for 5 different noise
levels and $3$ different data sizes.  We observe that the PSD features, i.e.
one of the most used technique for model discrimination in cosmology is unable
to perform the desired task. The main reason is that the two different model
have very similar PSDs (see \figref{psd_sigma3}).   As a result, one
need to use other statistics to solve the problem. In this case, we use
histogram features and observe a significant improvement over the PSD.  The
SCNN present the advantage to learn automatically specific features for the
task and also the usual Deep Learning performance bump.

@@

\section{Spherical harmonics VS graph Fourier eigenvectors}
\figref{g:graph_harmonics} displays the first 16 eigenvectors of the
graph Fourier basis used for the convolution. While their is a clear
resemblance with the spherical harmonic, the small irregularities in the
sampling (see \figref{healpix_graph_4}) have two important effects on the
graph Fourier eigenvectors.  First, some graph eigenvectors will be
localized~\cite{perraudin2018global}, i.e: they will span only a small subpart
of the sphere.  Second, as Nside increase toward infinity and hence the number
of pixel $n\rightarrow \infty$, we are still unsure if the graph eigenvectors
converge toward the spherical harmonics. We believe that it might be the case
for a different graph construction (fully connected) and that a proof could be
built on top of the work of~\cite{belkin2007convergence}. This is obviously
out of the scope of this contribution.

@@

\section{Filter visualization}
\label{sec:filter_visualization}

In \figref{gaussian_filters_visualization}, we present three different
ways to plot the graph convolution kernel $K_t(x)=e^{-\tau t x}$. First, they can
be plotted in the graph spectral domain. In this case, we simply evaluate $K_t$
at the graph eigenvalues $\text{diag}(\bLambda)$. Second, we can convolve a
delta on the sphere and plot a gnomonic view of the results. Third, since we are
working on the sphere, we can only plot the section of the convoluted delta. In
\figref{index_section}, we display the selected indexes for the section
plotting. Note that, because of the small irregularities in the HealPix
sampling, the second and the third methods are likely to change depending on the
chosen delta.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/index_plotting_order20_nside16.pdf}
\caption{Indexes selected for the section ploting of \figref{gaussian_filters_visualization} middle. The delta was placed on the yellow node.}
\label{fig:index_section}
\end{figure}

@@

When comparing the network, we want to ensure that the SVM classifier has
access to the same amount of data that the network, i.e, potentially an
infinite number of sample. Hence, we fit different linear SVM classifier using
various amount of training set size until we oberve experimentally that
increasing the amount of data does not improve the classifier. The validation
set remain the same and is used for cross-validation to tune the L2
penalization parameter of SVM. Eventually, we evaluate the final classifier on
the testing set. An example of the evolution of the training and valdication
error is show in \figref{hist_error_evolution}.
